Probabilistic forecasting of hourly Emergency Department arrivals

### 2023-04-13 journal_submission/revised/latex_without_author/paper_without_authors.tex
prepare a version to share in the journal

  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
    \author[CBS]{Bahman Rostami-Tabar%
  \corref{cor1}%
  }
   \ead{rostami-tabarb@cardiff.ac.uk} 
    \author[UG]{Jethro Browell%
  %
  }
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[LU]{Ivan Svetunkov%
  %
  }
   \ead{i.svetunkov@lancaster.ac.uk} 
      \affiliation[CBS]{Cardiff Business School, Cardiff University, UK.}
    \affiliation[UG]{School of Mathematics and Statistics, University of Glasgow, UK.}
    \affiliation[LU]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK.}
    \cortext[cor1]{Corresponding author}

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing (Svetunkov, 2021a), Autoregressive Integrated Moving Average
(ARIMA) (Hyndman and Athanasopoulos, 2021), Autoregressive Conditional
Heteroskedasticity (ARCH) (Bollerslev et al., 1994), Vector Autoregressive
model (Lütkepohl, 2013), TBATS (De Livera et al., 2011) and Artificial
Neural Networks (Hyndman and Athanasopoulos, 2021).We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below. Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals. There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e. having more staff than needed is costly, but
having less staff than required may lead to worse outcomes for patients. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient (Wright et al., 2006). Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently (Ramos et al., 2013; Rostami-Tabar and Ziel, 2022). Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. Another drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches. Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.

@@

There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention. We refer interested readers to some extensive reviews of the relevant literature by Shi et al. (2022), Gul and Celik (2020) and Ibrahim et al. (2016). The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare. However, given the focus of this study, we only discuss hourly ED forecasting.

@@

Figure \ref{fig:seasonplot-dofw} illustrates the daily subseries plot, with the x-axis representing the date and y-axis
the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
of week from Monday to Saturday. The blue line shows the average arrival for the given day.
It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
compared to the rest of the week. This might be due to the closure of General Practitioners
outpatient clinics over the weekends.

@@

Prophet is a forecasting procedure created by Facebook
(Taylor and Letham, 2018) that accounts for multiple seasonality,
piecewise trend and holiday effects. Prophet is robust to missing data
and shifts in the trend and typically handles outliers well. Prophet
works well on daily data seen in Facebook. It is robust and automated,
making it easy to learn for beginners. The implementation may be less
flexible than other methods. The model itself relies on the Multiple
Source of Error state space model, initially proposed by Kalman (1960).
The model is incorporated using the corresponding implementation of the
Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
package (O'Hara-Wild et al., 2020). Note that the input data is assigned with an
hourly and daily seasonality. This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
benchmarks.

@@

Probabilistic forecasts are evaluated following the principle of
\emph{sharpness subject to calibration}, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \ref{fig:quantile-bias},
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and \(t\) family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. Notably, both benchmarks exhibit nega-
tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
This could result in poor staffing decisions. This is because the empirical distribution of
whole data fails to characterise how arrivals in ED may change over time.

@@

\textbackslash The main benefit of the applied models is the use of probabilistic forecasts to inform decision making. Probabilistic forecasts contain all potential future outcomes and help planners to
achieve more efficient decisions by not only predicting the most likely outcome but also
quantifying the probability of all possible outcomes including extremely high or extremely
low arrivals in the Emergency Departments. This information enables decision makers to
manage risk associated with low-probability-high-impact events. Based on these forecasts,
hospitals can decide how many nurses to have for each shift to make the work of the ED more
efficient, e.g.~to meet the service level targets set by the National Health Service (NHS) in
the United Kingdom.

Practically speaking, the running time can be an important aspect for managers depending
on the frequency of generating forecasts. If it is high, one may employ models that do not
require a lot of computational time, such as Poisson regression (Poisson 1), sacrificing the
accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
like GAMLSS (Ttr-2) should be used.

Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning. Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.

@@

The dataset used in this study does not include the period of the COVID-19 pandemic.
During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
any forecasting model used for ED arrivals forecasting during that period would need to be
modified to reflect those changes for that specific period. One of the simplest modifications
would be to include a set of dummy variables, capturing different stages of the pandemic.
However, this is outside of the scope of this paper and can be considered as a direction for
future research.

>>>>

  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
      \cortext[cor1]{Corresponding author}

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing (Svetunkov, 2021a), Autoregressive Integrated Moving Average
(ARIMA) (Hyndman and Athanasopoulos, 2021), Autoregressive Conditional
Heteroskedasticity (ARCH) (Bollerslev et al., 1994), Vector Autoregressive
model (Lütkepohl, 2013), TBATS (De Livera et al., 2011) and Artificial
Neural Networks (Hyndman and Athanasopoulos, 2021). \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e. having more staff than needed is costly, but
having less staff than required may lead to worse outcomes for patients}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient (Wright et al., 2006). Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently (Ramos et al., 2013; Rostami-Tabar and Ziel, 2022). Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches. Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}

@@

\textcolor{blue}{There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention. We refer interested readers to some extensive reviews of the relevant literature by} Shi et al. (2022), Gul and Celik (2020) and Ibrahim et al. (2016). \textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare. However, given the focus of this study, we only discuss hourly ED forecasting.}

@@

Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
of week from Monday to Saturday. The blue line shows the average arrival for the given day.
It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
compared to the rest of the week. This might be due to the closure of General Practitioners
outpatient clinics over the weekends.}

@@

Prophet is a forecasting procedure created by Facebook
(Taylor and Letham, 2018) that accounts for multiple seasonality,
piecewise trend and holiday effects. Prophet is robust to missing data
and shifts in the trend and typically handles outliers well. Prophet
works well on daily data seen in Facebook. It is robust and automated,
making it easy to learn for beginners. The implementation may be less
flexible than other methods. The model itself relies on the Multiple
Source of Error state space model, initially proposed by Kalman (1960).
The model is incorporated using the corresponding implementation of the
Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
package (O'Hara-Wild et al., 2020). Note that the input data is assigned with an
hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
benchmarks.}

@@

Probabilistic forecasts are evaluated following the principle of
\emph{sharpness subject to calibration}, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \ref{fig:quantile-bias},
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and \(t\) family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks exhibit nega-
tive quantile bias as they struggle to capture the long term trend of increasing arrivals.
This could result in poor staffing decisions. This is because the empirical distribution of
whole data fails to characterise how arrivals in ED may change over time.}

@@

\textbackslash textcolor\{blue\}\{The main benefit of the applied models is the use of probabilistic forecasts to inform decision
making. Probabilistic forecasts contain all potential future outcomes and help planners to
achieve more efficient decisions by not only predicting the most likely outcome but also
quantifying the probability of all possible outcomes including extremely high or extremely
low arrivals in the Emergency Departments. This information enables decision makers to
manage risk associated with low-probability-high-impact events. Based on these forecasts,
hospitals can decide how many nurses to have for each shift to make the work of the ED more
efficient, e.g.~to meet the service level targets set by the National Health Service (NHS) in
the United Kingdom.

\textcolor{blue}{Practically speaking, the running time can be an important aspect for managers depending
on the frequency of generating forecasts. If it is high, one may employ models that do not
require a lot of computational time, such as Poisson regression (Poisson 1), sacrificing the
accuracy of forecasts. If the frequency is lower and running time is not a big concern, a model
like GAMLSS (Ttr-2) should be used.}

\textcolor{blue}{Probabilistic forecasts of hourly ED arrivals can benefit ED process because it provides the timing and the magnitude of the unlikely scenarios with huge impact on the service delivery, which are fundamental for capacity planning. Probabilistic forecasts can be used to better manage risks of under and over resource allocation, which consequently can reduce both costs and risks for patients, staff and the service as a whole.}

@@

\textcolor{blue}{
The dataset used in this study does not include the period of the COVID-19 pandemic.
During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
any forecasting model used for ED arrivals forecasting during that period would need to be
modified to reflect those changes for that specific period. One of the simplest modifications
would be to include a set of dummy variables, capturing different stages of the pandemic.
However, this is outside of the scope of this paper and can be considered as a direction for
future research.}

### 2023-04-13 manuscript/paper.Rmd
prepare a version to share in the journal

journal: "Health Systems"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float,lscape,amsmath,bm} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: natbib
bibliography: mybibfile.bib
#biblio-style: authoryear
#csl: elsevier-without-titles.csl
#biblio-style: abbrevnat
editor_options: 
  markdown: 
    wrap: 72
---

@@

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
[@derlet2002overcrowding]. Also, it is related to increasing length of
stay [@muhammet2015forecasting], low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records and other [@ROSTAMITABAR20221197].

@@

```{r seasonplot-dofw, fig.cap= "Subseries plot: day of weekly arrivals", fig.align='center',out.width="70%"}
```

@@

Finally, Figure \@ref(fig:date-plot) presents the time plot of daily
arrivals. Each point represents one day, and points are shape-coded by
day-of-week to show the weekly cycle. The figure shows more arrivals on
Mondays than on other weekdays, agreeing with the previous findings.
Moreover, we can observe a long-term trend line and the significant
effect of some holidays. We can see that arrivals near Christmas and New
Year's day are significantly lower than other days of the year.
Moreover, Figure \@ref(fig:date-plot) allows us to identify the impact
of special events on arrivals. For instance, we see that arrivals are
significantly low for 01-03 of March 2018. These days correspond to the
Storm Emma with heavy snowfall that resulted in travel disruption, mass
power outages, and schools closed in the UK [@stormemma2018].

>>>>

journal: "Health Systems"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float,lscape,amsmath,bm} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: default
bibliography: mybibfile.bib
biblio-style: apa
csl: elsevier-harvard.csl
#biblio-style: abbrevnat
editor_options: 
  markdown: 
    wrap: 72
---

@@

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
[@derlet2002overcrowding]. Also, it is related to increasing length of
stay [@muhammet2015forecasting], low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records [@ROSTAMITABAR20221197].

@@

Clear weekday effects
are notable with larger volumes of incidents observed on
Fridays and Saturdays. All such observations will become
of key importance when designing schedules for ambulance
crews.

@@

In summary, we observe an increase in demand levels for
EMS between 2005 and 2009, and note that demand is
more volatile on weekends and in months associated with
school holidays, such as July and December.

### 2023-04-13 manuscript/paper.tex
prepare a version to share in the journal

  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
      \cortext[cor1]{Corresponding author}

@@

{\centering \includegraphics[width=0.7\linewidth]{paper_files/figure-latex/seasonplot-dofw-1}

\caption{Subseries plot: day of weekly arrivals}\label{fig:seasonplot-dofw}
\end{figure}

@@

Finally, Figure \ref{fig:date-plot} presents the time plot of daily
arrivals. Each point represents one day, and points are shape-coded by
day-of-week to show the weekly cycle. The figure shows more arrivals on
Mondays than on other weekdays, agreeing with the previous findings.
Moreover, we can observe a long-term trend line and the significant
effect of some holidays. We can see that arrivals near Christmas and New
Year's day are significantly lower than other days of the year.
Moreover, Figure \ref{fig:date-plot} allows us to identify the impact
of special events on arrivals. For instance, we see that arrivals are
significantly low for 01-03 of March 2018. These days correspond to the
Storm Emma with heavy snowfall that resulted in travel disruption, mass
power outages, and schools closed in the UK (Wales, n.d.).

@@

\leavevmode\vadjust pre{\hypertarget{ref-stormemma2018}{}}%
Wales, B., n.d. Wales returning to normal as snow thaws and temperatures rise.

>>>>

  \title{Probabilistic forecasting of hourly Emergency Department arrivals}
    \author[CBS]{Bahman Rostami-Tabar%
  \corref{cor1}%
  }
   \ead{rostami-tabarb@cardiff.ac.uk} 
    \author[UG]{Jethro Browell%
  %
  }
   \ead{jethro.browell@glasgow.ac.uk} 
    \author[LU]{Ivan Svetunkov%
  %
  }
   \ead{i.svetunkov@lancaster.ac.uk} 
      \affiliation[CBS]{Cardiff Business School, Cardiff University, UK.}
    \affiliation[UG]{School of Mathematics and Statistics, University of Glasgow, UK.}
    \affiliation[LU]{Centre for Marketing Analytics and Forecasting, Lancaster University, UK.}
    \cortext[cor1]{Corresponding author}

@@

Clear weekday effects
are notable with larger volumes of incidents observed on
Fridays and Saturdays. All such observations will become
of key importance when designing schedules for ambulance
crews.

@@

In summary, we observe an increase in demand levels for
EMS between 2005 and 2009, and note that demand is
more volatile on weekends and in months associated with
school holidays, such as July and December.

@@
